---
title: 'Data Brief #2'
author: "Abbie Raza"
date: "26/11/2021"
output: html_document
---

# 1. Research Question and Importance

Is there a relationship between socioeconomic status and substance use during the COVID-19 global pandemic? ~~This is the research question I plan on exploring. In particular, I hoped to see if there~~ **To answer this question, I studied whether there** was a correlation between ~~factors like~~ employment status and education level, which are markers of socioeconomic status, and consumption of cannabis and alcohol. 

~~This question is important because current reports have shown that~~ Unemployment has skyrocketed due to the global pandemic (Hensher, 2020), and thus has an impact on the socioeconomic status of individuals within North America. According to ~the same report by~~ Hensher (2020), the extended economic problems caused by COVID-19 could lead to negative health outcomes such as increased substance misuse as a result of growing living costs, unemployment, and a lack of access to healthcare resources, all of which contribute to socioeconomic status. SES is ~~important because it is~~ a significant indicator explaining trends between substance use and SES (Phelan et.al, 2010). FTC *What does FTC stand for?* suggests that SES ~~has an impact on~~ **affects** access to a wide range of resources that protect health and avoid health inequities, both socially and economically. Higher socioeconomic status has the unique ability to increase access to resources required to engage in preventative strategies to avoid poor health and hazardous habits, as well as to address health issues once they arise (Phelan et. al, 2010).  Several studies have found the link between the two variablesvto be significant indicators for the other~~ (Singh et. al, 2018; Patrick et. al, 2012). However, the direction of causality ~~between the two variables~~ is unclear as there is no way to probabilistically discern whether substance use is dependent on SES or vice versa (Glei and Weinstein 2019; Nicholson Jr., 2020). Many studies have hypothesized that they both potentially contribute to the other. 

In general, the specific trends for substance use have shown that higher alcohol use is correlated with a higher SES, whereas substances like cannabis and opioids are correlated with a lower SES. In both cases, the type of substance is correlated strongly with the level of SES. Based on this previous research, we should expect to see consistent results. For those of higher socioeconomic status (i.e., those who hold a bachelor’s degree or higher and are employed), ~~we canv **I** expect to see higher rates of alcohol use. And for those with lower socioeconomic status (i.e. those who do not hold a bachelor’s degree and are unemployed), ~~we can~~ **I** expect to see higher rates of cannabis use. 

*There are a lot of opportunities to condense your writing. Notice the places where I have removed text.*

# 2.) Methods

-   necessary libraries

```{r setup, include=FALSE}
library(dplyr)
library(descr)
library(DescTools)
library(ggplot2)
library(survey)
```

-   necessary functions and previous code

```{r}
Mode <- function(x) {
a = table(x)	# x is a vector   # function to return mode value 
return(a[which.max(a)])
}


data <- read.csv("cpss-5311-E-2021-sus_F1.csv", header=TRUE) # original 
d1 <- subset(subset(data,select = c(PEDUC_LC, PEMPSTC, ALC_05,ALC_20,ALC_35,CAN_05,CAN_15,CAN_30)))
d1 <- na.omit(d1)

d1$BAPLUS <- rep(0,length(d1$PEDUC_LC))
d1$BAPLUS <- ifelse(d1$PEDUC_LC >= 6, "Bachelor's", "No Bachelor's")
table(d1$BAPLUS)

d1$EMPLOY <- rep(0,length(d1$PEMPSTC))
d1$EMPLOY[d1$PEMPSTC == 1] <- "Employed and at work"
d1$EMPLOY[d1$PEMPSTC == 2 | d1$PEMPSTC == 3] <- "Employed but absent from work"
d1$EMPLOY[d1$PEMPSTC == 4] <- "Not employed"

d1$EMPLOY <- factor(d1$EMPLOY, 
                       levels = c("Employed and at work", 
                                  "Employed but absent from work",
                              
                                  "Not employed"))
table(d1$EMPLOY)


d1$ALCFREQ <- rep(0,length(d1$ALC_05))
d1$ALCFREQ [d1$ALC_05 ==1] <- "Yes"
d1$ALCFREQ [d1$ALC_05 == 2] <- "No"
d1$ALCFREQ [d1$ALC_05 == 3] <- "Never Drank Alcohol"

d1$ALCFREQ <- factor(d1$ALCFREQ, 
                       levels = c("Yes",
                                  "No",
                                  "Never Drank Alcohol"))
table(d1$ALCFREQ)


d1$ALCCON <- rep(0,length(d1$ALC_20))
d1$ALCCON [d1$ALC_20 == 1] <- "Increased"
d1$ALCCON [d1$ALC_20 == 2] <- "Decreased"
d1$ALCCON [d1$ALC_20 == 4] <- "Same"
d1$ALCCON [d1$ALC_20 == 3 | d1$ALC_20 == 6] <- "Not Applicable/ Valid Skip"

d1$ALCCON <- factor(d1$ALCCON, 
                       levels = c("Increased",
                                  "Decreased",
                                  "Same", "Not Applicable/ Valid Skip"))
table(d1$ALCCON)

d1$ALCPROB <- rep(0,length(d1$ALC_35))
d1$ALCPROB [d1$ALC_35 == 1] <- "Yes"
d1$ALCPROB [d1$ALC_35 == 2] <- "No"
d1$ALCPROB [d1$ALC_35 == 6] <- "Valid Skip"

d1$ALCPROB <- factor(d1$ALCPROB, 
                       levels = c("Yes",
                                  "No",
                                  "Valid Skip"))
table(d1$ALCPROB)


d1$CANFREQ <- rep(0,length(d1$CAN_05))
d1$CANFREQ [d1$CAN_05 == 1] <- "Never used cannabis"
d1$CANFREQ [d1$CAN_05 == 2] <- "Not in the past 30 days"
d1$CANFREQ [d1$CAN_05 == 3 | d1$CAN_05 == 4] <- "At least 1-3 times in the past 30 days"
d1$CANFREQ [d1$CAN_05 >= 5 & d1$CAN_05 < 8] <- "At least once a week"
d1$CANFREQ [d1$CAN_05 == 8] <- "daily"

d1$CANFREQ <- factor(d1$CANFREQ, 
                       levels = c("Never used cannabis",
                                  "Not in the past 30 days",
                                  "At least 1-3 times in the past 30 days",
                                  "At least once a week",
                                  "daily"))
table(d1$CANFREQ)

# recode CAN_15 how has cannabis increased (CANCON)
d1$CANCON <- rep(0,length(d1$CAN_15))
d1$CANCON [d1$CAN_15 == 1] <- "Increased"
d1$CANCON [d1$CAN_15 == 2] <- "Decreased"
d1$CANCON [d1$CAN_15 == 4] <- "Same"
d1$CANCON [d1$CAN_15 == 3 | d1$CAN_15 == 6] <- "Not Applicable/ Valid Skip"

d1$CANCON <- factor(d1$CANCON, 
                       levels = c("Increased",
                                  "Decreased",
                                  "Same",
                                  "Not Applicable/ Valid Skip"))
table(d1$CANCON)

# recode CAN_30 to CANPROB

d1$CANPROB <- rep(0,length(d1$CAN_30))
d1$CANPROB [d1$CAN_30 == 1] <- "Yes"
d1$CANPROB [d1$CAN_30 == 2] <- "No"
d1$CANPROB [d1$CAN_30 == 6] <- "Valid Skip"

d1$CANPROB <- factor(d1$CANPROB, 
                       levels = c("Yes",
                                  "No",
                                  "Valid Skip"))
table(d1$CANPROB)
d1 <- na.omit(d1)
```

-   subsetting cannabis and alcohol into two separate data sets

```{r}
# cannabis 
cann <- subset(subset(d1, select = c(CANCON, CANPROB, EMPLOY, BAPLUS)))
cann <- subset(subset(cann, d1$CANCON != "Not Applicable/ Valid Skip" & d1$CANPROB != "Valid Skip"), na.rm = TRUE)


# alcohol 
alc <- subset(subset(d1, select = c(ALCCON, ALCPROB, EMPLOY, BAPLUS)))
alc <- subset(subset(alc, d1$ALCCON != "Not Applicable/ Valid Skip" & d1$ALCPROB != "Valid Skip"), na.rm = TRUE)
alc <- na.omit(alc)
```

Measures of Association- ALCFREQ and BAPLUS/EMPLOY - Does an association exist? - How strong is the association? - What is the pattern or direction of the association?

```{r}
round(prop.table(table(d1$ALCFREQ))*100,2)
round(prop.table(table(d1$ALCFREQ, d1$BAPLUS))*100,2)
round(prop.table(table(d1$ALCFREQ, d1$EMPLOY))*100,2)

# given that a person is employed, what is the probability that they also consumed alcohol?
round(prop.table(table(d1$ALCFREQ, d1$EMPLOY), margin = 1),2) # option 1
CrossTable(d1$ALCFREQ, d1$EMPLOY, 
            prop.c = FALSE, prop.t = TRUE,
            prop.r = TRUE, prop.chisq = FALSE, total.r = TRUE, 
            total.c = TRUE) # option 2


# given that a person has at least a bachelor's, what is the probability that they also drank alcohol?
round(prop.table(table(d1$ALCFREQ, d1$BAPLUS), margin = 1),2)
CrossTable(d1$ALCFREQ, d1$BAPLUS, 
            prop.c = FALSE, prop.t = TRUE,
            prop.r = TRUE, prop.chisq = FALSE, total.r = TRUE, 
            total.c = TRUE) # option 2


# --------- CHI SQUARED TEST ---------
# Null Hypothesis: H0: there is no significant relationship between consuming alcohol in the past 30 days and employment status 
chisq.test(d1$ALCFREQ, d1$EMPLOY)

# Null Hypothesis: H0: there is no significant relationship between consuming alcohol in the past 30 days and whether or not respondents hold at least a bachelor's degree 
chisq.test(d1$ALCFREQ, d1$BAPLUS)

# measure of associations- employment status and ALCFREQ
GoodmanKruskalGamma(d1$ALCFREQ, d1$EMPLOY)
# measure of association- education level and ALCFREQ
GoodmanKruskalGamma(d1$ALCFREQ, d1$BAPLUS)

# barplot- CANFREQ and EMPLOY 
ggplot(data = subset(d1, !is.na(ALCFREQ)), aes(x = EMPLOY, y = (..count..)/sum(..count..), fill = ALCFREQ)) + 
  geom_bar(position = "dodge") +
    labs(x = "Employment Status", 
       y = "Percentage of total", 
       title = "Consumption of Alcohol in The Past 30 Days") +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 1)) + coord_flip()

# barplot- CANFREQ and BAPLUS 
ggplot(data = subset(d1, !is.na(ALCFREQ)), aes(x = BAPLUS, y = (..count..)/sum(..count..), fill = ALCFREQ)) + 
  geom_bar(position = "dodge") +
    labs(x = "Employment Status", 
       y = "Percentage of total", 
       title = "Consumption of Alcohol in The Past 30 Days") +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 1)) + coord_flip()

```

-   Measures of Association (CANFREQ and EMPLOY/BAPLUS)

```{r}
round(prop.table(table(d1$CANFREQ))*100,2)
round(prop.table(table(d1$CANFREQ, d1$BAPLUS))*100,2)
round(prop.table(table(d1$CANFREQ, d1$EMPLOY))*100,2)

# given that a person is employed, what is the probability that they also consumed cannabis?
CrossTable(d1$CANFREQ, d1$EMPLOY, 
            prop.c = FALSE, prop.t = TRUE,
            prop.r = TRUE, prop.chisq = FALSE, total.r = TRUE, 
            total.c = TRUE)


# given that a person has at least a bachelor's, what is the probability that they also consumed cannabis?
CrossTable(d1$CANFREQ, d1$BAPLUS, 
            prop.c = FALSE, prop.t = TRUE,
            prop.r = TRUE, prop.chisq = FALSE, total.r = TRUE, 
            total.c = TRUE) 


# --------- CHI SQUARED TEST ---------
# Null Hypothesis: H0: there is no significant relationship between consuming cannabis in the past 30 days and employment status 
chisq.test(d1$CANFREQ, d1$EMPLOY)

# Null Hypothesis: H0: there is no significant relationship between consuming cannabis in the past 30 days and whether or not respondents hold at least a bachelor's degree 
chisq.test(d1$CANFREQ, d1$BAPLUS)

# measure of associations- employment status and CANFREQ
GoodmanKruskalGamma(d1$CANFREQ, d1$EMPLOY)
# measure of association- education level and CANFREQ
GoodmanKruskalGamma(d1$CANFREQ, d1$BAPLUS)

# barplots 
# barplot- CANFREQ and EMPLOY 
ggplot(data = d1, aes(x = EMPLOY, y = (..count..)/sum(..count..), fill = CANFREQ)) + 
  geom_bar(position = "dodge") +
    labs(x = "Employment Status", 
       y = "Percentage of total", 
       title = "Consumption of Cannabis in The Past 30 Days") +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 1)) + coord_flip()
# barplot- CANFREQ and BAPLUS 
ggplot(data = d1, aes(x = BAPLUS, y = (..count..)/sum(..count..), fill = CANFREQ)) + 
  geom_bar(position = "dodge") +
    labs(x = "Employment Status", 
       y = "Percentage of total", 
       title = "Consumption of Cannabis in The Past 30 Days") +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 1)) + coord_flip()
```

-   Measures of Association (Cannabis only)

-   CANCON

```{r}
round(prop.table(table(cann$CANCON))*100,2)
round(prop.table(table(cann$CANCON, cann$EMPLOY))*100, 2)
round(prop.table(table(cann$CANCON, cann$BAPLUS))*100, 2)
# --- CANCON
# given that a person is employed, how likely is it that they consume cannabis?
CrossTable(cann$CANCON, cann$EMPLOY, 
           prop.c = FALSE, prop.t = TRUE,
           prop.r = TRUE, prop.chisq = FALSE, total.r = TRUE, 
           total.c = TRUE) 

# given that a person holds at least a bachelor's degree, how likely is it that they consume cannabis?
CrossTable(cann$CANCON, cann$BAPLUS, 
           prop.c = FALSE, prop.t = TRUE,
           prop.r = TRUE, prop.chisq = FALSE, total.r = TRUE, 
           total.c = TRUE) 

# --------- CHI SQUARED TEST ---------
# Null Hypothesis: H0: there is no significant relationship frequency of cannabis consumption in the past 30 days and employment status 
chisq.test(cann$CANCON, cann$EMPLOY)

# Null Hypothesis: H0: there is no significant relationship between frequency of cannabis consumption in the past 30 days and whether or not respondents hold at least a bachelor's degree 
chisq.test(cann$CANCON, cann$BAPLUS)

# measure of associations- employment status and CANCON
GoodmanKruskalGamma(cann$CANCON, cann$EMPLOY)
# measure of association- education level and CANCON
GoodmanKruskalGamma(cann$CANCON, cann$BAPLUS)

# barplot- CANCON and EMPLOY 
ggplot(data = subset(cann, !is.na(CANCON)), aes(x = EMPLOY, y = (..count..)/sum(..count..), fill = CANCON)) + 
  geom_bar(position = "dodge") +
  labs(x = "Employment Status", 
       y = "Percentage of total", 
       title = "Frequency of Consumption of Cannabis in The Past 30 Days") +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 1)) + coord_flip()

# barplot- CANCON and BAPLUS 
ggplot(data = subset(cann, !is.na(CANCON)), aes(x = BAPLUS, y = (..count..)/sum(..count..), fill = CANCON)) + 
  geom_bar(position = "dodge") +
  labs(x = "Employment Status", 
       y = "Percentage of total", 
       title = "Frequency of Consumption of Cannabis in The Past 30 Days") +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 1)) + coord_flip()

```

-   CANPROB

```{r}
round(prop.table(table(cann$CANPROB))*100,2)
round(prop.table(table(cann$CANPROB, cann$EMPLOY))*100, 2)
round(prop.table(table(cann$CANPROB, cann$BAPLUS))*100, 2)
# --- CANPROB
# given that a person is employed, how likely is it that they consume cannabis knowing that they had a problem?
CrossTable(cann$CANPROB, cann$EMPLOY, 
           prop.c = FALSE, prop.t = TRUE,
           prop.r = TRUE, prop.chisq = FALSE, total.r = TRUE, 
           total.c = TRUE) 

# given that a person holds at least a bachelor's degree, how likely is it that they consume cannabis knowing that they have a problem?
CrossTable(cann$CANPROB, cann$BAPLUS, 
           prop.c = FALSE, prop.t = TRUE,
           prop.r = TRUE, prop.chisq = FALSE, total.r = TRUE, 
           total.c = TRUE) 

# --------- CHI SQUARED TEST ---------
# Null Hypothesis: H0: there is no significant relationship cannabis consumption problems in the past 30 days and employment status 
chisq.test(cann$CANPROB, cann$EMPLOY)
fisher.test(cann$CANPROB, cann$EMPLOY)
# Null Hypothesis: H0: there is no significant relationship between cannabis consumption problems in the past 30 days and whether or not respondents hold at least a bachelor's degree 
chisq.test(cann$CANPROB, cann$BAPLUS)
fisher.test(cann$CANPROB, cann$BAPLUS)

# measure of associations- employment status and CANPROB
GoodmanKruskalGamma(cann$CANPROB, cann$EMPLOY)
# measure of association- education level and CANPROB
GoodmanKruskalGamma(cann$CANPROB, cann$BAPLUS)

# barplot- CANPROB and EMPLOY 
ggplot(data = subset(cann, !is.na(CANPROB)), aes(x = EMPLOY, y = (..count..)/sum(..count..), fill = CANPROB)) + 
  geom_bar(position = "dodge") +
  labs(x = "Employment Status", 
       y = "Percentage of total", 
       title = "Consumed Cannabis while knowing you had a problem\nas a function of employment status") +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 1)) + coord_flip()

# barplot- CANPROB and BAPLUS 
ggplot(data = subset(cann, !is.na(CANPROB)), aes(x = BAPLUS, y = (..count..)/sum(..count..), fill = CANPROB)) + 
  geom_bar(position = "dodge") +
  labs(x = "Employment Status", 
       y = "Percentage of total", 
       title = "Consumed cannabis while knowing that you had a problem\nas a function of education level") +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 1)) + coord_flip()
```

-   Measures of Association (Alcohol only)

-   ALCPROB

```{r}
round(prop.table(table(alc$ALCPROB))*100,2)
round(prop.table(table(alc$ALCPROB, alc$EMPLOY))*100, 2)
round(prop.table(table(alc$ALCPROB, alc$BAPLUS))*100, 2)
# --- ALCPROB
# given that a person is employed, how likely is it that they consume alcohol knowing that they had a problem?
CrossTable(alc$ALCPROB, alc$EMPLOY, 
           prop.c = FALSE, prop.t = TRUE,
           prop.r = TRUE, prop.chisq = FALSE, total.r = TRUE, 
           total.c = TRUE) 

# given that a person holds at least a bachelor's degree, how likely is it that they consume alcohol knowing that they have a problem?
CrossTable(alc$ALCPROB, alc$BAPLUS, 
           prop.c = FALSE, prop.t = TRUE,
           prop.r = TRUE, prop.chisq = FALSE, total.r = TRUE, 
           total.c = TRUE) 

# --------- CHI SQUARED TEST ---------
# Null Hypothesis: H0: there is no significant relationship alcohol consumption problems in the past 30 days and employment status 
chisq.test(alc$ALCPROB, alc$EMPLOY)

# Null Hypothesis: H0: there is no significant relationship between alcohol consumption problems in the past 30 days and whether or not respondents hold at least a bachelor's degree 
chisq.test(alc$ALCPROB, alc$BAPLUS)

# measure of associations- employment status and ALCPROB
GoodmanKruskalGamma(alc$ALCPROB, alc$EMPLOY)
# measure of association- education level and ALCPROB
GoodmanKruskalGamma(alc$ALCPROB, alc$BAPLUS)

# barplot- ALCPROB and EMPLOY 
ggplot(data = subset(alc, !is.na(ALCPROB)), aes(x = EMPLOY, y = (..count..)/sum(..count..), fill = ALCPROB)) + 
  geom_bar(position = "dodge") +
  labs(x = "Employment Status", 
       y = "Percentage of total", 
       title = "Consumed alcohol knowing that you had a problem\nas a function of employment status") +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 1)) + coord_flip()

# barplot- ALCPROB and BAPLUS 
ggplot(data = subset(alc, !is.na(ALCPROB)), aes(x = BAPLUS, y = (..count..)/sum(..count..), fill = ALCPROB)) + 
  geom_bar(position = "dodge") +
  labs(x = "Employment Status", 
       y = "Percentage of total", 
       title = "Consumed alcohol knowing that you had a problem\nas a function of education level") +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 1)) + coord_flip()
```

-   ALCCON

```{r}
round(prop.table(table(alc$ALCCON))*100,2)
round(prop.table(table(alc$ALCCON, alc$EMPLOY))*100, 2)
round(prop.table(table(alc$ALCCON, alc$BAPLUS))*100, 2)
# --- ALCCON
# given that a person is employed, how likely is it that they consumed alcohol in the past 30 days?
CrossTable(alc$ALCCON, alc$EMPLOY, 
           prop.c = FALSE, prop.t = TRUE,
           prop.r = TRUE, prop.chisq = FALSE, total.r = TRUE, 
           total.c = TRUE) 
# given that a person consumed alcohol in the past 30 days, how likely is that they are employed?
CrossTable(alc$ALCCON, alc$EMPLOY, 
           prop.c = TRUE, prop.t = TRUE,
           prop.r = FALSE, prop.chisq = FALSE, total.r = TRUE, 
           total.c = TRUE) 

# given that a person holds at least a bachelor's degree, how likely is it that they consume alcohol in the past 30 days?
CrossTable(alc$ALCCON, alc$BAPLUS, 
           prop.c = FALSE, prop.t = TRUE,
           prop.r = TRUE, prop.chisq = FALSE, total.r = TRUE, 
           total.c = TRUE) 

# given that a person consumed alcohol in the past 30 days, how likely is it that they hold a bachelor's degree?
CrossTable(alc$ALCCON, alc$BAPLUS, 
           prop.c = TRUE, prop.t = TRUE,
           prop.r = FALSE, prop.chisq = FALSE, total.r = TRUE, 
           total.c = TRUE)
# --------- CHI SQUARED TEST ---------
# Null Hypothesis: H0: there is no significant relationship alcohol consumption in the past 30 days and employment status 
chisq.test(alc$ALCCON, alc$EMPLOY)

# Null Hypothesis: H0: there is no significant relationship between alcohol consumption in the past 30 days and whether or not respondents hold at least a bachelor's degree 
chisq.test(alc$ALCCON, alc$BAPLUS)

# measure of associations- employment status and ALCCON
GoodmanKruskalGamma(alc$ALCCON, alc$EMPLOY)
# measure of association- education level and ALCCON
GoodmanKruskalGamma(alc$ALCCON, alc$BAPLUS)

# barplot- ALCCON and EMPLOY 
ggplot(data = subset(alc, !is.na(ALCCON)), aes(x = EMPLOY, y = (..count..)/sum(..count..), fill = ALCCON)) + 
  geom_bar(position = "dodge") +
  labs(x = "Employment Status", 
       y = "Percentage of total", 
       title = "Consumed alcohol in the past 30 days as a function of employment status") +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 1)) + coord_flip()

# barplot- ALCCON and BAPLUS 
ggplot(data = subset(alc, !is.na(ALCCON)), aes(x = BAPLUS, y = (..count..)/sum(..count..), fill = ALCCON)) + 
  geom_bar(position = "dodge") +
  labs(x = "Employment Status", 
       y = "Percentage of total", 
       title = "Consumed alcohol in the past 30 days as a function of education level") +
  theme(axis.text.x = element_text(angle = 0, 
                                   hjust = 1)) + coord_flip()
```
To test for significance, I used a chi-squared test of significance. Finally, to measure the strength of the association between the dependent and independent variables, I used Goodman-Kruskal Gamma. 

ALCFREQ and EMPLOY/BAPLUS

Starting with ALCFREQ, which is the frequency of alcohol consumption in the past 30 days, roughly 68.8% of the respondents answered yes. 39.81% did not hold a bachelor’s degree, compared to those who did hold a bachelor’s degree at 28.97%. When it came to employment status, 42.8% of those who were currently employed reported consuming alcohol. These results are consistent with the conditional bar plots. 
When comparing cross-tabulated data, given that a person is employed, they were more likely to consume alcohol at least once in the past 30 days compared to those who were absent from work and those who were currently unemployed. Of those who were employed (both at work or absent from work), employed respondents were more likely to consume alcohol compared to unemployed respondents. Following this, those who held (at least) a bachelor’s degree were less likely to consume alcohol than those without a bachelor’s degree in the past 30 days. 
For the Chi-squared test between ALCFREQ and EMPLOY, the p-value is significant at a p <0.05 confidence interval. In other words, the association between the frequency of alcohol consumption and their employment status is significant 95% of the time. When it came to ALCFREQ and BAPLUS, the p-value is significant at p < 0.05. So the association between frequency of alcohol consumption and education level is significant, a result we can be sure of 95% of the time. 
Finally, when looking at the Goodman-Kruskal Gamma, the strength of association between ALCFREQ and EMPLOY was weak at roughly 0.15. The association between ALCFREQ and BAPLUS was negligible at roughly 0.094. 
 

CANFREQ and EMPLOY/BAPLUS 

CANFREQ, which is the frequency of cannabis consumption in the past 30 days, a good majority reported that they had not. However, given a respondent’s employment status, those who were employed were also less likely to use cannabis at least once in the past 30 days. For BAPLUS, those who held no bachelor's degree were more likely to use cannabis at least once in the past 30 days.
In the chi-squared test between CANFREQ and EMPLOY, the p-value is significant at a p < 0.05 confidence interval. In other words, the association between the frequency of cannabis consumption and their employment status is significant 95% of the time. When it came to CANFREQ and BAPLUS, the p-value is significant at a p < 0.05 confidence interval. So the association between frequency of cannabis consumption and education level is a significant result we can be sure of 95% of the time. 
Finally, when looking at the Goodman-Kruskal gamma, the strength of association between CANFREQ and EMPLOY was -0.24, indicating a moderate inverse relationship. Thus, those who were employed were less likely to use cannabis in the past 30 days, generally speaking. When it came to education level, the  Goodman Kruskal gamma was 0.007 indicating a negligible correlation between the two variables, BAPLUS and CANFREQ. 
 

CANCON and EMPLOY/BAPLUS 

Of the respondents who claimed to have used cannabis in the past 30 days, those who were employed claimed that their cannabis consumption had remained relatively the same for the most part. The same can be said for those who are unemployed as well. When it came to education level, those who held at least a bachelor's degree reported that their cannabis consumption remained the same for the most part, however, the increase in cannabis use was nearly the same proportion to those who claim that their cannabis consumption had remained the same. There was a much larger gap between unemployed respondents, where a larger majority claimed that their cannabis consumption remained the same. 
The chi-squared test between CANCON and EMPLOY yielded an insignificant p-value at p < 0.05. Therefore, we fail to reject the null hypothesis that there is no significant relationship between cannabis use and employment status. In other words, the association between the consumption of cannabis use trends and employment status is insignificant 95% of the time. For the same confidence interval, the relationship between BAPLUS and CANCON is significant at p < 0.05.
The Goodman-Kruskal Gamma between employment status and CANCON was 0.10, indicating a weak relationship between the two variables. The same can be said between education level and CANCON at 0.15, which is slightly higher than the relationship between CANCON and employment status, but still weak. 
 

CANPROB and EMPLOY/BAPLUS

Of The respondents who claimed to have used cannabis in the past 30 days, The majority of respondents claimed that they had not consumed cannabis knowing that they had a problem. Of the 61 people who reported that they consume cannabis while knowing that they had a problem, 43 of them were employed and at work and 5 of them were employed but absent from work, comprising 78.7% of those consuming cannabis while knowing that they had a problem- however, this proportion can also be explained by the number of people who were employed at work being a lot larger than those who were unemployed.
The chi-squared test between CANPROB and EMPLOY yielded an insignificant p-value at p < 0.05. Therefore, we fail to reject the null hypothesis that there is no significant relationship between cannabis use problems and employment status. In other words, the association between the consumption of cannabis problem trends and employment status is insignificant 95% of the time. This was further confirmed by Fisher’s exact test. For BAPLUS and CANPROB, the chi-squared test also yielded an insignificant result at p < 0.05. Thus, the relationship between BAPLUS and CANCON is more likely due to random chance; we can be sure of this 95% of the time. 
The Goodman Kruskal Gamma for CANPROB and EMPLOY is weak at 1.5 and even weaker for education level at 0.04. 
 

ALCCON and EMPLOY/BAPLUS 

Of the respondents who responded yes to alcohol consumption (ALCFREQ) in the past 30 days, the majority of them responded to having the same level of consumption. However, when looking at the bar plot, the difference between the consumption of alcohol during the pandemic between those who were employed versus those who were unemployed, there were a couple differences. One, the proportion of employed people at work whose alcohol consumption increased was larger compared to the proportion of employed individuals at work whose alcohol consumption decreased. The opposite is true for unemployed respondents. Unemployed respondents saw a larger proportional decrease in alcohol consumption compared to an increase. Two, those who were employed but absent from work shared consistent trends with those who were employed, in that even individuals who were absent from work but still employed reported a noticed increase in alcohol use compared to a decrease. As for the level of education, there was a noticeable increase in both groups for alcohol consumption compared to a decrease. However, for the majority of individuals, alcohol consumption remained the same regardless of education level. 
The Chi-squared test for ALCCON and EMPLOY showed a significant p-value at p < 0.05. Therefore, we can reject the null hypothesis and conclude that there is a significant relationship between alcohol consumption and employment status, and we can be sure of this at the 95% confidence interval. The same is true for the relationship between ALCCON and BAPLUS. Therefore, we can conclude that there is a significant relationship between employment status and ALCCON and between education level and ALCCON. 
Goodman-Kruskal Gamma for ALCCON and EMPLOY was 0.13, showing a weak relationship between the two variables. For ALCCON and BAPLUS, this relationship was even weaker at 0.079. 
 

ALCPROB and EMPLOY/BAPLUS

For both employment status and education level, the vast majority of respondents answered no, meaning that they had not consumed alcohol knowing that they had a problem. 
The Chi-squared test for ALCPROB and EMPLOY yielded a significant p-value at p <0.05. Therefore, we can reject the null hypothesis and conclude that there is a significant relationship between ALCPROB and EMPLOY, and we can be sure of this at the 95% confidence interval. However, the Chi-squared test for ALCPROB and BAPLUS was insignificant at p < 0.05. Therefore, we fail to reject the null hypothesis and conclude that there is no significant relationship between ALCPROB and BAPLUS. 
The Goodman-Kruskal Gamma measure of association for ALCPROB and EMPLOY was 0.19, indicating a weak to a moderate relationship. For ALCPROB and BAPLUS, the relationship was an inverse weak relationship. Those who had a higher level of education were less likely to consume alcohol knowing that they had a problem. However, this relationship is relatively weak. 
 


# 3.) Analysis and Results

multivariate/ logistic regression between cannabis, employment, and education

- CANFREQ, ALCFREQ and EMPLOY/BAPLUS 

```{r}

# recoded into logic variables 
d1$CANFREQ2 <- rep(0,length(d1$CANFREQ))
d1$CANFREQ2 [d1$CANFREQ == "Never used cannabis" | 
               d1$CANFREQ == "Not in the past 30 days" ] <- 1
# d1$CANFREQ2[d1$CANFREQ == "At least 1-3 times in the past 30 days"] <- 0 
# d1$CANFREQ2[d1$CANFREQ == "At least once a week" | d1$CANFREQ == "daily"] <- 0
table(d1$CANFREQ2)

d1$ALCFREQ2 <- rep(0,length(d1$ALCFREQ))
d1$ALCFREQ2 [d1$ALCFREQ == "No" | d1$ALCFREQ == "Never Drank Alcohol"] <- 1
table(d1$ALCFREQ2)


# logistic regression for CANFREQ
regCANFREQ2 <- glm(CANFREQ2 ~ EMPLOY + BAPLUS, family = binomial(link = logit), data = d1) 
summary(regCANFREQ2)
PseudoR2(regCANFREQ2)

# logistic regression for ALCFREQ 
regALCFREQ <- glm(ALCFREQ2 ~ EMPLOY + BAPLUS, family = binomial(link = logit), data = d1) 
summary(regALCFREQ)
PseudoR2(regALCFREQ)

```


- Logistic Regression for Cannabis (CANCON and CANPROB)
```{r}
cann$CANCON2 <- rep(0,length(cann$CANCON))
cann$CANCON2[cann$CANCON == "Increased" | cann$CANCON == "Same"] <- 1
cann$CANCON2[cann$CANCON == "Decreased"] <- 0
table(cann$CANCON2)

cann$CANPROB2 <- rep(0,length(cann$CANPROB))
cann$CANPROB2[cann$CANPROB == "Yes"] <- 1
cann$CANPROB2[cann$CANPROB == "No"] <- 0
table(cann$CANPROB2)

regCANCON <- glm(CANCON2 ~ EMPLOY + BAPLUS, family = binomial(link = logit), data = cann) 
summary(regCANCON)
# McFadden's pseudo r-squared 
PseudoR2(regCANCON)


regCANPROB <- glm(CANPROB2 ~ EMPLOY + BAPLUS, family = binomial(link = logit), data = cann) 
summary(regCANPROB)
# McFadden's pseudo r-squared
PseudoR2(regCANPROB)
```




multivariate/ logistic regression between alcohol, employment, and education

```{r}
alc$ALCCON2 <- rep(0,length(alc$ALCCON))
alc$ALCCON2[alc$ALCCON == "Increased" | alc$ALCCON == "Same"] <- 1
alc$ALCCON2[alc$ALCCON == "Decreased"] <- 0
table(alc$ALCCON2)

alc$ALCPROB2 <- rep(0,length(alc$ALCPROB))
alc$ALCPROB2[alc$ALCPROB == "Yes"] <- 1
alc$ALCPROB2[alc$ALCPROB == "No"] <- 0
table(alc$ALCPROB2)


regALCCON <- glm(ALCCON2 ~ EMPLOY + BAPLUS, family = binomial(link = logit), data = alc) 
summary(regALCCON)
# McFadden's pseudo r-squared 
PseudoR2(regALCCON)

regALCPROB <- glm(ALCPROB2 ~ EMPLOY + BAPLUS, family = binomial(link = logit), data = alc) 
summary(regALCPROB)
# McFadden's pseudo r-squared 
PseudoR2(regALCPROB)
```

Since all of the variables in my dataset are ordinal categorical variables, I used logistic regression to see the weighted effects of employment status and education level when it came to looking at consumption trends for cannabis and alcohol. 
 
CANFREQ and EMPLOY/ BAPLUS 
Some of the key findings for cannabis use were that overall, employment and education level were significant predictors for cannabis use. There was a positive estimated coefficient between CANFREQ and unemployment. Additionally, there was a negative log odds correlation between having a bachelor’s degree and cannabis use. The pseudo-r-squared for this finding is 0.02, showing a weak correlation and indicating that there could be other factors contributing to cannabis use patterns. In other words, while these factors are significant predictors, they are not the only variables contributing to cannabis use. 
 
ALCFREQ and EMPLOY/BAPLUS 
Overall, employment status and education levels were significant predictors for alcohol use, where not being employed had a positive log-odds estimate of 0.600 and having no bachelor’s has a negative log-odds estimate of -0.29. Again, the pseudo-r-squared for this relationship was very weak as well, at 0.02, roughly speaking. In other words, while these factors are significant predictors, they are not the only variables contributing to alcohol use. 
 
CANCON, CANPROB, and EMPLOY/BAPLUS 
Overall, education level was a significant predictor for CANCON, where having no bachelor’s had a positive log odds of 0.53. The pseudo-r-squared for this relationship was 0.012, indicating a weak correlation and that there are other factors to consider that could be contributing to CANCON. 
For CANPROB, there were no significant relationships between employment, education level, and whether someone decided to use cannabis knowing that they had a problem. The pseudo-r-squared for this relationship is 0.006, indicating that it is not very likely that these two variables are good predictors for determining whether someone decided to use cannabis knowing that they have a problem that is made worse with cannabis. 
 
ALCCON, ALCPROB, and EMPLOY/BAPLUS 
Alcohol consumption had no significant predictors for employment status and education level. The pseudo-r-squared for ALCCON and EMPLOY/BAPLUS was 0.0009, indicating that employment status and education levels were that it is not very likely that these two variables are good predictors for determining the level of alcohol consumed when compared together. 
As for ALCPROB, both employment and education were significant predictors for determining whether or not someone consumed alcohol knowing that they had a problem made worse by alcohol. being employed but absent from work and not being employed were both significant. However, those who were unemployed had a negative log-odds estimate, and those who were employed and were absent from work had a positive log-odds estimate. Additionally, those without a bachelor’s degree had a positive log-odds estimate with ALCPROB. The pseudo-r-squared for this regression was 0.011, indicating a weak correlation, where there are other factors correlating to ALCPROB. 
 
For my findings, some of the limitations are definitely having only 2 predictor variables. In reality, SES is comprised of multiple factors besides education level and employment status, like familial wealth, job status, income, geographical location, etc. So the fact that there are significant correlations between substance use and employment/education but that these correlations are weak is not surprising. Additionally, the sampling method used for this study excluded people who lived and worked in remote areas like indigenous reserves, which would skew the results and would not speak for all Canadians within the labour force. Keeping these limitations in mind, in summary, it seems that employment and education are significant predictors for substance use, however, they are not the primary predictors and are most likely linked to other factors like income, wealth, health status among many other variables. 







References: 

Glei, D. A., & Weinstein, M. (2019). Drug and alcohol abuse: the role of economic insecurity. American journal of health behavior, 43(4), 838-853.

Hensher, M. (2020). Covid-19, unemployment, and health: time for deeper solutions?. BMJ, 371.

Nicholson Jr, H. L. (2020). Socioeconomic status, fundamental cause theory, and prescription opioid use behaviors: a theoretical examination. Sociological Spectrum, 40(1), 1-32.

Patrick, M. E., Wightman, P., Schoeni, R. F., & Schulenberg, J. E. (2012). Socioeconomic status and substance use among young adults: a comparison across constructs and drugs. Journal of studies on alcohol and drugs, 73(5), 772-782.

Phelan, J. C., & Link, B. G. (2013). Fundamental cause theory. In Medical sociology on the move (pp. 105-125). Springer, Dordrecht.

Singh, S., Martins, D., Khuu, W., Tadrous, M., Gomes, T., & Juurlink, D. (2018). Characteristics of Opioid-Related Deaths in Ontario, Canada: Leveraging the Drug and Drug/Alcohol Related Death (DDARD) Database. International Journal of Population Data Science, 3(4).